Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
Fix figure 3.
fixed the formatting of lens expressions

The expression \bcp{mangled---did you proofread this?}\cd{project [
  \n\t\r]+ -> " "} is a QRE that matches any 
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{attheedges}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}}
  \label{fig:attheedges}
\end{figure}

@@

\begin{verbatim}
let bib_to_end : (lens in bibtex <=> endnot) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{verbatim}
\noindent to synthesize a quotient lens that maps between \verb!bibtex! and
\verb!endnote! QREs, and sends the equivalence class of \verb!bib_bexample! to
the equivalence class of \verb!end_example!.

@@

We have already encountered the \bcp{Again, formatting is mangled.  Also,
  are we setting lens expressions in typewriter font, or in snas-serif?  I
  don't care, but we need to be consistent.}\verb!project R $\to$ s! QRE
which sends all elements of the language described by the regular expression
$R$, and the

@@

The equivalence described is one in which there is a bijection between 
the regular expressions \verb!comma_name! and \verb!space_name! given by,
\begin{verbatim}
let comma_name : regexp = name | (name . ", " . name . (" " . name)*)

@@

\end{verbatim}

\noindent with the lens \verb!comma_to_space! mapping between them given by 
\begin{verbatim}
let comma_to_space : (lens in space_name <=> comma_name) =
lens_swap 
(name . del " " . (ins " " . name . del " ")* ) (name . ins ", ") || name 
\end{verbatim}

@@

\begin{verbatim}
let squash_author : canonizer =  
squash space_name $\to$ comma_name using (get comma_to_space)

let author : canonizer = "author = \"" . squash_author . "\""
\end{verbatim}

In this example, the \verb!squash_author! QRE expresses the fact that the
\verb!space_name! and \verb!comma_name! representations of the authors' name
are equivalent, with the \verb!comma_name! representation chosen as the
canonical representation, and the function \verb!comma_to_space.get! giving
the canonizing function which converts data in the \verb!space_name! format
to its \verb!comma_name! representation.

@@

\begin{verbatim}
let comma_to_space : (lens in space_name <=> comma_name) = 
synth space_name <=> comma_name using {("J. H. Conway", "Conway, J. H.")}
\end{verbatim}

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex{} record is named
\verb!bibtex!, then the QRE for matching two records is 
\verb!bibtex . bibtex!, the QRE for matching one or two records is
\verb! bibtex | (bibtex . bibtex)! and the QRE for matching any number of QRE
records \verb!bibtex*!

@@

\begin{verbatim}
let ambiguous : canonizer = "a"* . (project "a"* $\to$ "a")
\end{verbatim}

\noindent The behaviour of \verb!ambiguous! is not-well defined since the string
\verb!"aaa"! can be canonized to any of \verb!"a", "aa", "aaa"! depending on how
\verb!"aaa"! is parsed. We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.

@@

The more interesting combinators are the $\mathit{lquot}$ and $\mathit{rquot}$
combinators. The $\mathit{lquot}(c, q)$ combinator takes a quotient lens $q$ and
quotients the source data using $c$, assuming that the source data forms a
complete set of representatives for the equivalence relation $\eqrel{c}$. The
$\mathit{lquot}(q, c)$ combinator does the same but on the view data. For
example, recall that in the \bibtex{} to EndNote transformation, we had the QREs
\verb!bibtex! and \verb!endnote! describing the \bibtex{} and EndNote records
respectively, and the bijective lens \verb!bib_to_end! that maps between
\verb!bibtex! and \verb!endnote!. The quotient lens which maps between
\verb!bibtex! and \verb!endnote! is given by the QRE lens

\begin{verbatim}
let bib_to_end_q : (lens in bibtex <=> endnote) = 
rquot (lquot (bibtex, bib_to_end), endnote)
 \end{verbatim}

@@

More specifically, in order to synthesize a quotient lens $q: c \Leftrightarrow
c'$, that is consistent with the examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$
 it suffices to perform the following procedure:
\begin{enumerate}
  \item
  Compute $K(c)$, $K(c')$,
  \item
  compute $\{({x_1}', {y_1}'), \ldots, ({x_n}', {y_n}')\}$ where ${x_i}' =
  \canonizer(c)(x_i)$ and ${y_i}' = \canonizer(c')(y_i)$,
  \item
    synthesize a bijective lens $\ell : K(c) \Leftrightarrow K(c')$ that is
    consistent with the examples $\{({x_1}', {y_1}'), \ldots, ({x_n}',
    {y_n}')\}$
  \item 
  return $q = \mathit{rquot}(\mathit{lquot}(c, \ell), c') : c \Leftrightarrow
  c'$
\end{enumerate}
As an example, recall that
in the \bibtex{} to EndNote transformation we had the QREs \verb!bibtex! and \verb!endnote! describing a \bibtex{} record and an
Endnote record respectively. We also had the example pair 
\verb!(bib_example , end_example)! where

\begin{verbatim}
bib_example = 
"@Book {conway,
  Author = \"J. H. Conway\",
  Title = {Regular Algebra and Finite Machines},
}"
\end{verbatim}

@@

Since the transformation between the kernel of \verb!bibtex! and the kernel of
\verb!endnote! that maps \verb!bib_example! to \verb!end_example! can be
expressed by a bijective lens, it suffices to type the expression

\begin{verbatim}
let bib_to_end : (lens in bibtex <=> endnot) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{verbatim}

>>>>

The expression \cd{project [\n\t\r]}$^+$ $\to$ \cd{" "} is a QRE that matches any 
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{canonizers-outside}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}\bcp{Done---what do you think?}}
  \label{fig:attheedges}
\end{figure}

@@

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnot) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
\noindent to synthesize a quotient lens that maps between \cd{bibtex} and
\cd{endnote} QREs, and sends the equivalence class of \cd{bib_bexample} to
the equivalence class of \cd{end_example}.

@@

We have already encountered the \cd{project R} $\to$ \cd{s} QRE
which sends all elements of the language described by the regular expression
$R$, and the

@@

The equivalence described is one in which there is a bijection between 
the regular expressions \cd{comma_name} and \cd{space_name} given by,
\begin{lstlisting}
let comma_name : regexp = name | (name . ", " . name . (" " . name)*)

@@

\end{lstlisting}

\noindent with the lens \cd{comma_to_space} mapping between them given by 
\begin{lstlisting}
let comma_to_space : (lens in space_name <=> comma_name) =
lens_swap 
(name . del " " . (ins " " . name . del " ")* ) (name . ins ", ") || name 
\end{lstlisting}

@@

\begin{lstlisting}
let squash_author : canonizer =  
squash space_name $\to$ comma_name using (get comma_to_space)

let author : canonizer = "author = \"" . squash_author . "\""
\end{lstlisting}

In this example, the \cd{squash_author} QRE expresses the fact that the
\cd{space_name} and \cd{comma_name} representations of the authors' name
are equivalent, with the \cd{comma_name} representation chosen as the
canonical representation, and the function \cd{comma_to_space.get} giving
the canonizing function which converts data in the \cd{space_name} format
to its \cd{comma_name} representation.

@@

\begin{lstlisting}
let comma_to_space : (lens in space_name <=> comma_name) = 
synth space_name <=> comma_name using {("J. H. Conway", "Conway, J. H.")}
\end{lstlisting}

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex{} record is named
\cd{bibtex}, then the QRE for matching two records is 
\cd{bibtex . bibtex}, the QRE for matching one or two records is
\cd{ bibtex | (bibtex . bibtex)} and the QRE for matching any number of QRE
records \cd{bibtex*}

@@

\begin{lstlisting}
let ambiguous : canonizer = "a"* . (project "a"* $\to$ "a")
\end{lstlisting}

\noindent The behaviour of \cd{ambiguous} is not-well defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a", "aa", "aaa"} depending on how
\cd{"aaa"} is parsed. We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.

@@

The more interesting combinators are the $\mathit{lquot}$ and $\mathit{rquot}$
combinators. The $\mathit{lquot}(c, q)$ combinator takes a quotient lens $q$ and
quotients the source data using $c$, assuming that the source data forms a
complete set of representatives for the equivalence relation $\eqrel{c}$. The
$\mathit{lquot}(q, c)$ combinator does the same but on the view data. For
example, recall that in the \bibtex{} to EndNote transformation, we had the QREs
\cd{bibtex} and \cd{endnote} describing the \bibtex{} and EndNote records
respectively, and the bijective lens \cd{bib_to_end} that maps between
\cd{bibtex} and \cd{endnote}. The quotient lens which maps between
\cd{bibtex} and \cd{endnote} is given by the QRE lens

\begin{lstlisting}
let bib_to_end_q : (lens in bibtex <=> endnote) = 
rquot (lquot (bibtex, bib_to_end), endnote)
 \end{lstlisting}

@@

More specifically, in order to synthesize a quotient lens $q: c \Leftrightarrow
c'$, that is consistent with the examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$
 it suffices to perform the following procedure:
\begin{enumerate}
  \item
  Compute $K(c)$, $K(c')$,
  \item
  compute $\{({x_1}', {y_1}'), \ldots, ({x_n}', {y_n}')\}$ where ${x_i}' =
  \canonizer(c)(x_i)$ and ${y_i}' = \canonizer(c')(y_i)$,
  \item
    synthesize a bijective lens $\ell : K(c) \Leftrightarrow K(c')$ that is
    consistent with the examples $\{({x_1}', {y_1}'), \ldots, ({x_n}',
    {y_n}')\}$
  \item 
  return $q = \mathit{rquot}(\mathit{lquot}(c, \ell), c') : c \Leftrightarrow
  c'$
\end{enumerate}
As an example, recall that
in the \bibtex{} to EndNote transformation we had the QREs \cd{bibtex} and \cd{endnote} describing a \bibtex{} record and an
Endnote record respectively. We also had the example pair 
\cd{(bib_example , end_example)} where

\begin{lstlisting}
bib_example = 
"@Book {conway,
  Author = \"J. H. Conway\",
  Title = {Regular Algebra and Finite Machines},
}"
\end{verbatim}

@@

Since the transformation between the kernel of \cd{bibtex} and the kernel of
\cd{endnote! that maps \cd{bib_example! to \cd{end_example} can be
expressed by a bijective lens, it suffices to type the expression

\begin{lstlisting}
let bib_to_end : (lens in bibtex <=> endnot) 
  = synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{lstlisting}
